---
title: "miRNA target genes"
output: html_document
date: "2025-09-16"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = 1) # Warnings are printed immediately, not delayed (some packages delay warning printing).
```

```{r}
library(dplyr)
library(ggplot2)
library(ggpubr)
source("~/Source/miRNA/miRNA_functions.R")

library(org.Hs.eg.db)
```

```{r}
setwd("/mnt/Data1/HwangLab/Arun/Adult_Freeze-040625/")
```

# 0. Annotation

miRNA mapping
```{r}
load("0_Annot/miRNA_index.Rdata")
miRge.id.expanded = count.id.expanded; rm(count.id.expanded)
miRge.index = index.annot; rm(index.annot) # miRge alignment index
miRge.index.seq = index.fa; rm(index.fa)
miRBase.coord = hsa.gff.pair; rm(hsa.gff.pair)
```

Target DB
```{r}
mirdb = readRDS("/mnt/Data1/HwangLab/Arun/Adult_Freeze-040625/Annotation/miRNA_annot_mirdb.rds")
dim(mirdb)
```

> score ≥ 80: High confidence that the target is real, score < 60: Requires caution and additional supporting evidence. 

# 1. DE

```{r}
load("~/HwangLab/taeyoungh/miRNA/Results/freeze040625_DE.Rdata")
```

### mRNA background

```{r}
gencode.gene.table = readRDS("~/HwangLab/taeyoungh/miRNA/Adult_Freeze-082825/gencode.gene.table.rds")
```

```{r}
mirdb = mutate(mirdb, de.dlpfc.expressed = gene_id %in% subset(gencode.gene.table, dlpfc_cauc.expressed & dlpfc_aa.expressed)$gene_id,
               de.mpfc.expressed = gene_id %in% subset(gencode.gene.table, mpfc_cauc.expressed & mpfc_aa.expressed)$gene_id,
               de.hippo.expressed = gene_id %in% subset(gencode.gene.table, hippo_cauc.expressed & hippo_aa.expressed)$gene_id,
               de.caudate.expressed = gene_id %in% subset(gencode.gene.table, caudate_cauc.expressed & caudate_aa.expressed)$gene_id)

mirdb = mutate(mirdb, de.pfc.expressed = gene_id %in% subset(gencode.gene.table, dlpfc_cauc.expressed & dlpfc_aa.expressed & mpfc_cauc.expressed & mpfc_aa.expressed)$gene_id)
```

### Region DE

Link to miRBase
```{r}
# control background
miRBase.coord$de.region.all = miRgeToMirBase(rownames(de.region.res), miRBase.coord$miRNA) # 3 are not mapped finally: hsa-miR-1973, hsa-miR-339, hsa-miR-378g
miRBase.coord$de.region.sig = miRgeToMirBase(rownames(de.region.sig), miRBase.coord$miRNA)

# region specific
miRBase.coord$de.region.pfcSpecific = miRgeToMirBase(rownames(subset(de.region.sig, pfcSpecific>0.7)), miRBase.coord$miRNA)
miRBase.coord$de.region.hippoSpecific = miRgeToMirBase(rownames(subset(de.region.sig, hippoSpecific>0.7)), miRBase.coord$miRNA)
miRBase.coord$de.region.caudateSpecific = miRgeToMirBase(rownames(subset(de.region.sig, caudateSpecific>0.7)), miRBase.coord$miRNA)

miRBase.coord$de.region.pfcSpecific2 = miRgeToMirBase(rownames(subset(de.region.sig, pfcSpecific2>1)), miRBase.coord$miRNA)
miRBase.coord$de.region.hippoSpecific2 = miRgeToMirBase(c("hsa-miR-184", "hsa-miR-1911-5p", "hsa-miR-204-5p", "hsa-miR-4423-5p"), miRBase.coord$miRNA)
miRBase.coord$de.region.caudateSpecific2 = miRgeToMirBase(rownames(subset(de.region.sig, caudateSpecific2>1)), miRBase.coord$miRNA)
```

Target identification
```{r}
mirdb = mutate(mirdb, de.region.all = mir_id %in% subset(miRBase.coord, de.region.all!="Not")$miRNA,
               de.region.pfcSpecific = mir_id %in% subset(miRBase.coord, de.region.pfcSpecific!="Not")$miRNA,
               de.region.hippoSpecific = mir_id %in% subset(miRBase.coord, de.region.hippoSpecific!="Not")$miRNA,
               de.region.caudateSpecific = mir_id %in% subset(miRBase.coord, de.region.caudateSpecific!="Not")$miRNA)

mirdb = mutate(mirdb, de.region.pfcSpecific2 = mir_id %in% subset(miRBase.coord, de.region.pfcSpecific2!="Not")$miRNA,
               de.region.hippoSpecific2 = mir_id %in% subset(miRBase.coord, de.region.hippoSpecific2!="Not")$miRNA,
               de.region.caudateSpecific2 = mir_id %in% subset(miRBase.coord, de.region.caudateSpecific2!="Not")$miRNA)
```

GO
```{r}
de.region.go = list()

de.region.go[["all"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.region.all & (de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed) & score>80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.region.go[["all"]])
```

```{r}
de.region.go[["pfcSpecific2"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.pfc.expressed & de.region.pfcSpecific2 & score>80)$gene_id2), universe = unique(subset(mirdb, de.pfc.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.region.go[["pfcSpecific2"]])

de.region.go[["hippoSpecific2"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.hippo.expressed & de.region.hippoSpecific2 & score>80)$gene_id2), universe = unique(subset(mirdb, de.hippo.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.region.go[["hippoSpecific2"]])

de.region.go[["caudateSpecific2"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.caudate.expressed & de.region.caudateSpecific2 & score>80)$gene_id2), universe = unique(subset(mirdb, de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.region.go[["caudateSpecific2"]])
```

> caudate is close to subventricular zone (SVZ)...

### Race DE

Link to miRBase
```{r}
miRBase.coord$de.race.all = miRgeToMirBase(subset(de.race.res, region=="all")$mirgeID, miRBase.coord$miRNA)
miRBase.coord$de.race.sig = miRgeToMirBase(subset(de.race.sig, region=="all")$mirgeID, miRBase.coord$miRNA)

miRBase.coord$de.race.up = miRgeToMirBase(subset(de.race.sig, region=="all" & log2FoldChange>0.2)$mirgeID, miRBase.coord$miRNA) # aa > cauc
miRBase.coord$de.race.down = miRgeToMirBase(subset(de.race.sig, region=="all" & log2FoldChange<(-0.2))$mirgeID, miRBase.coord$miRNA) # aa < cauc
```

Target identification
```{r}
#mirdb = mutate(mirdb, de.race.dlpfc = mir_id %in% subset(miRBase.coord, de.race.dlpfc!="Not")$miRNA,
#               de.race.mpfc = mir_id %in% subset(miRBase.coord, de.race.mpfc!="Not")$miRNA,
#               de.race.hippo = mir_id %in% subset(miRBase.coord, de.race.hippo!="Not")$miRNA,
#               de.race.caudate = mir_id %in% subset(miRBase.coord, de.race.caudate!="Not")$miRNA)

mirdb = mutate(mirdb, de.race.up = mir_id %in% subset(miRBase.coord, de.race.up!="Not")$miRNA,
               de.race.down = mir_id %in% subset(miRBase.coord, de.race.down!="Not")$miRNA)
```

GO
```{r}
de.race.go = list()
# de.race.go[["dlpfc"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.dlpfc.expressed & de.race.dlpfc & score>=80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
# 
# clusterProfiler::dotplot(de.race.go[["dlpfc"]])
# 
# de.race.go[["caudate"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.caudate.expressed & de.race.caudate & score>=80)$gene_id2), universe = unique(subset(mirdb, de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
# 
# clusterProfiler::dotplot(de.race.go[["caudate"]])
# 
# de.race.go[["hippo"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.hippo.expressed & de.race.hippo & score>=80)$gene_id2), universe = unique(subset(mirdb, de.hippo.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
# 
# clusterProfiler::dotplot(de.race.go[["hippo"]])
# 
# de.race.go[["mpfc"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.mpfc.expressed & de.race.mpfc & score>=80)$gene_id2), universe = unique(subset(mirdb, de.mpfc.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
# 
# clusterProfiler::dotplot(de.race.go[["mpfc"]])
```

```{r}
de.race.go[["up"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed & de.race.up & score>=80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.race.go[["up"]])

de.race.go[["down"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed & de.race.down & score>=80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)

clusterProfiler::dotplot(de.race.go[["down"]])
```

### Significant terms
```{r}
de.regionSpecific.go.sig = list()
for (region in c("pfcSpecific2", "hippoSpecific2", "caudateSpecific2")) {
  de.regionSpecific.go.sig[[region]] = de.region.go[[region]]@result %>% mutate(geneRatio = parse_ratio(GeneRatio)) %>% dplyr::select(ID, Description, qvalue, geneRatio) %>% filter(qvalue<0.05)
}

de.regionSpecific.go.sig = do.call(rbind, de.regionSpecific.go.sig) 
de.regionSpecific.go.sig = de.regionSpecific.go.sig %>% tibble::rownames_to_column("temp") %>% mutate(region=stringr::word(temp, 1, sep = "\\.")) %>% dplyr::select(region, everything()) %>% dplyr::select(-temp)

with(de.regionSpecific.go.sig, table(region))
summary(de.regionSpecific.go.sig$geneRatio)
```

```{r}
de.race.go.sig = list()
for (region in c("up", "down")) {
  de.race.go.sig[[region]] = de.race.go[[region]]@result %>% mutate(geneRatio = parse_ratio(GeneRatio)) %>% dplyr::select(ID, Description, qvalue, geneRatio) %>% filter(qvalue<0.05)
}

de.race.go.sig = do.call(rbind, de.race.go.sig) 
de.race.go.sig = de.race.go.sig %>% tibble::rownames_to_column("temp") %>% mutate(region=stringr::word(temp, 1, sep = "\\.")) %>% dplyr::select(region, everything()) %>% dplyr::select(-temp)

with(de.race.go.sig, table(region))
summary(de.race.go.sig$geneRatio)
```

qvalue Plot
```{r}
de.regionSpecific.go.sig.qval = dcast(subset(de.regionSpecific.go.sig, geneRatio>0.05), Description~region, value.var = "qvalue") %>% tibble::column_to_rownames("Description")

de.regionSpecific.go.sig.qval = -log10(de.regionSpecific.go.sig.qval)

# all
temp = de.regionSpecific.go.sig.qval
temp[is.na(temp)] = -1
pheatmap::pheatmap(temp)

# common pathways
temp = de.regionSpecific.go.sig.qval[rowSums(is.na(de.regionSpecific.go.sig.qval))==0,]
dim(temp)
pheatmap::pheatmap(temp)

# region-specific
temp = de.regionSpecific.go.sig.qval[rowSums(is.na(de.regionSpecific.go.sig.qval))>0 & rowSums(is.na(de.regionSpecific.go.sig.qval))<3 ,]
temp[is.na(temp)] = -1
pheatmap::pheatmap(temp)
```

```{r}
de.race.go.sig.qval = dcast(subset(de.race.go.sig, geneRatio>0.05), Description~region, value.var = "qvalue") %>% tibble::column_to_rownames("Description")

de.race.go.sig.qval = -log10(de.race.go.sig.qval)

# all
temp = de.race.go.sig.qval
temp[is.na(temp)] = -1
pheatmap::pheatmap(temp)

# common pathways
temp = de.race.go.sig.qval[rowSums(is.na(de.race.go.sig.qval))==0,]
dim(temp)
pheatmap::pheatmap(temp)
```

# 2. eQTL

```{r}
load("~/HwangLab/taeyoungh/miRNA/Results/freeze082825_eQTL.Rdata")

miRqtl.mash.cauc.sig.large = miRqtl.mash.cauc.sig[apply(abs(miRqtl.mash.res[miRqtl.mash.cauc.sig, c("dlpfc.cauc_beta", "mpfc.cauc_beta", "hippo.cauc_beta", "caudate.cauc_beta")]), 1, max)>0.5]
miRqtl.mash.aa.sig.large = miRqtl.mash.aa.sig[apply(abs(miRqtl.mash.res[miRqtl.mash.aa.sig, c("dlpfc.aa_beta", "mpfc.aa_beta", "hippo.aa_beta", "caudate.aa_beta")]), 1, max)>0.5]
```

Link to miRBase
```{r}
miRBase.coord$miRqtl.mash.cauc.sig.large = miRgeToMirBase(sub(".*_", "", miRqtl.mash.cauc.sig.large), miRBase.coord$miRNA)
miRBase.coord$miRqtl.mash.aa.sig.large = miRgeToMirBase(sub(".*_", "", miRqtl.mash.aa.sig.large), miRBase.coord$miRNA)
```

Target identification
```{r}
mirdb = mutate(mirdb, 
               miRqtl.mash.cauc.sig.large = mir_id %in% miRBase.coord$miRqtl.mash.cauc.sig.large,
               miRqtl.mash.aa.sig.large = mir_id %in% miRBase.coord$miRqtl.mash.aa.sig.large)
```

GO
```{r}
miRqtl.go = list()

miRqtl.go[["cauc"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed & miRqtl.mash.cauc.sig.large & score>=80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
 
clusterProfiler::dotplot(miRqtl.go[["cauc"]])

miRqtl.go[["aa"]] = clusterProfiler::enrichGO(gene = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed & miRqtl.mash.aa.sig.large & score>=80)$gene_id2), universe = unique(subset(mirdb, de.dlpfc.expressed & de.mpfc.expressed & de.hippo.expressed & de.caudate.expressed)$gene_id2), keyType = "ENSEMBL", OrgDb = org.Hs.eg.db, ont = "BP", pAdjustMethod = "BH", qvalueCutoff = 0.05, readable = TRUE)
 
clusterProfiler::dotplot(miRqtl.go[["aa"]])
```

# Save
```{r}
save(miRBase.coord, mirdb, de.region.go, de.race.go, miRqtl.go, file="~/HwangLab/taeyoungh/miRNA/Results/freeze040625_target.Rdata")
```



